---
title: "Air_pollution_case_study"
author: "Bruce Okallau"
date: "6/17/2017"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy = TRUE)
```

# Air Pollution Case Study
This is my attempt at an exercise from Roger D Peng's "Explority Data Analysis" course on Coursera. Instead of comparing 1999 to 2012 I will compare 2012 to 2016.

## Step 1 - Read in the data
```{r}
all.data2012 <- read.csv("daily_aqi_by_cbsa_2012.csv")
head(all.data2012,20)
```

Note, this is slightly different from the example in the video due to changes in the EPA's data collection and warehousing. In order to compare only PM2.5 values I will need to create a subset.

```{r}
pm2012 <- subset(all.data2012, Defining.Parameter=="PM2.5")
head(pm2012)
```

Note: "sample value" is now called "AQI".

```{r}
x2012 <- pm2012$AQI
class(x2012)
```

```{r}
str(x2012)
summary(x2012)
```

The 2012 data available in this dataset is about half the size of the file in the example. There are no missing values.

```{r}
# performing the same steps for 2016
all.data2016 <- read.csv("daily_aqi_by_cbsa_2016.csv")
pm2016 <- subset(all.data2016, Defining.Parameter=="PM2.5")
x2016 <- pm2016$AQI
str(x2016)
summary(x2016)
```

There are comperable observations in 2016. The difference is that AQI are integers and the sample values are floats.

```{r}
#visualize the difference in the means, using log transform
boxplot(log10(x2012),log10(x2016))
```

## Exploring change at one monitor
```{r}
# does site 10140 have enough observations to be a good site?
nrow(subset(pm2012, CBSA.Code==10140))
nrow(subset(pm2016, CBSA.Code==10140))
```

```{r}
#subset for site 10140 for each year
site0 <- (subset(pm2012, CBSA.Code==10140))
site1 <- subset(pm2016, CBSA.Code==10140)
dim(site0)
dim(site1)
```

```{r}
# compare site observations over time first with 2012
plot(site0$Date, site0$AQI)
```

```{r}
# compare site observations over time now with 2016
plot(site1$Date, site1$AQI)
```

```{r}
rng <- range(site0$AQI, site1$AQI)
par(mfrow = c(1,2), mar= c(4,4,2,1))
plot(site0$Date, site0$AQI, pch= 20, ylim = rng)
abline(h = median(site0$AQI))
plot(site1$Date, site1$AQI, pch = 20, ylim = rng)
abline(h = median(site1$AQI))
```

The median AQI was higher in 2016 but not by much and the spread of AQI values was slightly smaller in 2016.

## How have States changed from 2012 to 2016
```{r}
# state is last two characters of CBSA string
pm2012$state <- str_sub(pm2012$CBSA, start= -2)
pm2016$state <- str_sub(pm2016$CBSA, start= -2)
```

```{r}
# what's the mean by state in 2012
mn0 <- with(pm2012, tapply(AQI, state, mean))
summary(mn0)
```

```{r}
# now for 2016
mn1 <- with(pm2016, tapply(AQI, state, mean))
#create dataframe with means and states
d0 <- data.frame(state = names(mn0), mean = mn0)
d1 <- data.frame(state = names(mn1), mean = mn1)
mrg <- merge(d0, d1, by = "state")
```

```{r}
# plot the state averages
par(mfrow = c(1,1))
rng1 <- range(mrg$mean.x, mrg$mean.y)
with(mrg, plot(rep(2012, 50), mrg[,2], xlim= c(2011, 2017),ylim=rng1))
with(mrg, points(rep(2016, 50), mrg[,3]))
segments(rep(2012, 50), mrg[,2], rep(2016, 50), mrg[,3])
```

We see mostly a downward trend.

```{r}
# states with highest means in 2012
highest.states.d0 <- d0[order(-d0$mean),]
head(highest.states.d0, 10)
```

```{r}
# states with highest means in 2016
highest.states.d1 <- d1[order(-d1$mean),]
head(highest.states.d1, 10)
```

```{r}
# states with loweest means in 2012
lowest.states.d0 <- d0[order(d0$mean),]
head(lowest.states.d0, 10)
```

```{r}
# states with loweest means in 2016
lowest.states.d1 <- d1[order(d1$mean),]
head(lowest.states.d1, 10)
```